---
title: 'The Landscape of Renal Replacement Therapy in Nigeria \| 2022 - 2023'
subtitle: 'A project of the renal registry committee of the Nigerian Association of Nephrologists'
author: 'Analysis by: Nelson Udeme-Abasi'
format: 'html'
execute:
  eval: true
  echo: true
  output: true
  warning: false
  error: false
  cache: true
---

# Analysis

## Summary Statistics

Loading necessary packages and loading the data

```{r}
# Loading the necessary packages
library(tidyverse)
library(gtsummary)
library(gt)
library(ggplot2)
library(survival)
library(survminer)
library(sf)
library(ggspatial)
library(prettymapr)
library(labelled)

# Loading the data
rr_data <- read_csv('rr_data.csv', trim_ws = T)
```

-   Selecting the necessary columns for my analysis and enforcing data types

-   Adding necessary columns

-   Forcing data types to ease analysis

```{r}
rr_data_new <- rr_data %>% 
  select(c(Sex, Age, Zone, State, `Centre Category`, `First Modality`, `Vascular Access`, DM, HBV, HIV, HCV, Outcome, `Date of Diagnosis`, `Date of Death or Transplant`))

# Creating labels for my table columns
col_labels <- c('Sex', 'Age(years)','Geopolitical Zone', 'State', 'Facility Type', 'First Modality','Vascular Access', 'DM Status', 'HBV Status', 'HIV Status', 'HCV Status', 'Outcomes', 'Date Diagnosed', 'Date of Death or Transplant')

rr_data_new <- set_variable_labels(.data = rr_data_new, .labels = col_labels)

rr_data_new <- rr_data_new %>%
  mutate('Age Group' = case_when(Age < 20 ~ 'Less than 20',
                                 Age >= 20 & Age <= 45 ~ '20 to 45',
                                 Age > 45 & Age <=65 ~ '46 to 65',
                                 Age > 65 ~ 'Greater than 65')
         ) %>% 
  mutate_at(c('Date of Diagnosis','Date of Death or Transplant'), mdy) %>% 
  mutate_at(c('Sex', 'Age Group', 'Zone', 'Centre Category', 'First Modality', 'Vascular Access', 'DM', 'HBV', 'HCV', 'HIV', 'Outcome'), as.factor)

head(rr_data_new)
```

Re-level factor variables to make sense of the data

```{r}
rr_data_new <- rr_data_new %>% 
  mutate(`Age Group` = fct_relevel(`Age Group`, 'Less than 20','20 to 45','46 to 65', 'Greater than 65'), 
         DM = fct_relevel(DM, 'Yes', 'No'),
         HIV = fct_relevel(HIV, 'Pos', 'Neg'),
         HBV = fct_relevel(HBV, 'Pos', 'Neg'),
         HCV = fct_relevel(HCV, 'Pos', 'Neg'),
         Outcome = fct_relevel(Outcome, 'Dead', 'Alive', 'Transplant')
         )
```

Below is the summary table of the patient's demographics

```{r}
#tbl_summary(rr_data_new,
# include = c(Sex, Age, `Age Group`, Zone, `Centre Category`),
#  statistic = list(
#    all_continuous() ~ '{median} ({p25}, {p75})',
#    all_categorical() ~ '{n} ({p}%)'
#  ),
#  digits = list(
#    all_categorical() ~ c(0, 1),
#    all_continuous() ~ c(1,1)
#  )
#  ) %>% 
#  modify_header(label ~ '**Variables**') %>% 
#  modify_caption('**Table 1 Sociodemographic Characteristics of Patients**') %>% 
#  bold_labels() %>% 
#  as_gt() %>% 
#  cols_width(everything() ~ px(250))
#  )
```

## Geographical Distribution of Patients (using the UN human data Nigerian shape file)

```{r}
# Bring in the UN human data Nigerian shape file and merge/join with the relevant summary in the data

nigeria <- read_sf('nga_admbnda_adm1_osgof_20190417.shp')

# Creating a summary of all states with patient's number

pts_geog_distr <- rr_data_new %>% 
  group_by(State) %>% 
  summarise(pts_per_state = n()) %>% 
  ungroup()

head(pts_geog_distr)
```

```{r}
# Confirm columns in State are similar to ADM1_EN

anti_join(pts_geog_distr, nigeria, by = c('State' = 'ADM1_EN'))

# Only FCT is not named in the same way

# So we will rename FCT in the pts_geog_distr df
pts_geog_distr <- pts_geog_distr %>% 
  mutate(State = case_when(State == 'FCT' ~ 'Federal Capital Territory',
                           TRUE ~ State)
         )
pts_geog_distr
```

```{r}
# Now safely combine both with a left join

map_data <- left_join(nigeria, pts_geog_distr, by = c('ADM1_EN' = 'State'))
```

Below is the map of `Nigeria` showing the state-level geographical distribution of patients who underwent RRT in 2022 - 2023

```{r}

#compute cenntroids for states label
state_centroids <- map_data %>%
  mutate(centroid = st_centroid(geometry)) %>%
  mutate(long = st_coordinates(centroid)[,1],
         lat  = st_coordinates(centroid)[,2])

# Now plot the map
map_data %>% 
  ggplot(aes(fill = pts_per_state)) +
  geom_sf()+
  geom_point(data = state_centroids, aes(x = long, y = lat))+
  geom_text(data = state_centroids, aes(x = long, y = lat, label = ADM1_EN),
            size = 3, color = "black", nudge_y = -.2) +
  scale_fill_gradient(
    low = "#d8f3dc", high = "#00A676", na.value = "grey90")+
  labs(
    title = "Geographical distribution of patients by State",
    caption = "Nigerian shapefile: gotten from humandata.org"
  ) +
  theme_bw()
```

## Patient's Distribution by Facility Type and Zone

Below is the patient distribution by facility category and zone

```{r}

rr_data_new %>% 
  group_by(Zone, `Centre Category`) %>% 
  summarise('No of Patients' = n()) %>% 
    ggplot(aes(x = Zone, y = `No of Patients`, fill = `Centre Category`, width = .7))+
      geom_bar(stat = 'identity', position = position_dodge(width = 0.75, preserve = "single"))+
      scale_fill_manual(values = c("Private" = "#E97132", "Public" = "#00A676")) +
      labs(title = '')+
      theme_bw()
```

## Age Distribution by Zone and Category

```{r}
ggplot(data = rr_data_new, aes(x = Zone, y = Age, fill = `Centre Category`))+
  stat_boxplot(geom = "errorbar",
               position = position_dodge(width = 0.75, preserve = "single"),
               width = 0.2
               )+
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 40, ymax = 60, 
           fill = "#FF0000", alpha = 0.3) +
  geom_boxplot(position = position_dodge(width = 0.75, preserve = "single"), width = .6)+
  scale_fill_manual(values = c("Private" = "#E97132", "Public" = "#00A676"))+
  geom_hline(yintercept = 60, linetype = 'dashed')+
  geom_hline(yintercept = 40, linetype = 'dashed')+
    theme_bw()
```

## Patients Distribution by Sex and Zone

```{r}
rr_data_new %>% 
  ggplot(aes(x = Zone, fill = Sex))+
  geom_bar(position = 'dodge')+
  scale_fill_manual(values = c("F" = "#E97132", "M" = "#00A676"))+
  geom_text(stat = "count", aes(label = after_stat(count)),
            position = position_dodge(width = 0.9),
            vjust = -0.5)+
  theme_bw()
```

## Temporal Trends in Kidney Failure Pattern

Below shows the visualization of the temporal trends in RRT in Nigeria (2022 - 2023)

```{r}
# Aggregrate dataframe to show monthly trends

rr_data_new %>% 
  mutate(monthly_rrt = floor_date(`Date of Diagnosis`, unit = 'months')) %>%
  group_by(monthly_rrt) %>% 
  summarise(monthly_rrt_pts = n()) %>% 
  filter(monthly_rrt >= ymd('2022-01-01') & monthly_rrt < ymd('2024-01-01')) %>% 
  ggplot()+
  geom_vline(xintercept = ymd('2022-03-01'), width = .1, color = 'red',
             linetype = 'dashed')+
  geom_vline(xintercept = ymd('2022-09-01'), width = .1, color = 'red',
             linetype = 'dashed')+
  geom_vline(xintercept = ymd('2023-03-01'), width = .1, color = 'red',
             linetype = 'dashed')+
  geom_vline(xintercept = ymd('2023-09-01'), width = .1, color = 'red',
             linetype = 'dashed')+
  annotate('text',x = ymd('2022-06-01'), y = 650, label = 'Dry Season', size = 3.5)+
  annotate('text',x = ymd('2022-12-01'), y = 650, label = 'Wet Season', size = 3.5)+
  annotate('text',x = ymd('2023-06-01'), y = 650, label = 'Dry Season', size = 3.5)+
  geom_line(aes(x = monthly_rrt, y = monthly_rrt_pts), color = "#00A676", width = 2)+
  labs(title = 'Temporal trends in Renal Failure Incidence',
       x = 'Monthly Trend',
       y= 'No of Patients')+
  theme_bw()
```

## Vascular Access (Overall and by Zone)

```{r}
# Overall Vascular Access
rr_data_new %>%
  filter(`Vascular Access` %in% c('AV Fistula','Temporary Femoral Catheter',
                                  'Tunneled CVC')) %>% 
  select(`Vascular Access`) %>% 
  group_by(`Vascular Access`) %>% 
  summarise(count = n(),
         `%Vascular Access` = paste0(round((count/4850)*100, 1),'%'))


```

```{r}
rr_data_new %>% 
  filter(`Vascular Access` %in% c('AV Fistula','Temporary Femoral Catheter',
                                  'Tunneled CVC')) %>%
  ggplot(aes(x = Zone, fill = `Vascular Access`)) +
  geom_bar(position = 'fill')+
  scale_fill_manual(values = c("AV Fistula" = "#E97132",
                              "Temporary Femoral Catheter" = "#00A676",
                              "Tunneled CVC" = "#FFC000"))+
  labs(title = 'Vascular Access by Zone',
       y = 'Percentage of total in zone')+
  theme_bw()
```

## Health Outcomes (by Zone)

```{r}
rr_data_new %>% 
  filter(Outcome %in% c('Dead','Transplant','Alive')) %>%
  ggplot(aes(y = Zone, fill = Outcome)) +
  geom_bar(position = 'fill')+
  scale_fill_manual(values = c("Dead" = "#E97132",
                              "Alive" = "#00A676",
                              "Transplant" = "#FFC000"))+
  labs(title = 'Outcomes by Zone',
       x = 'Percentage of total in zone')+
  theme_bw()
```

## Survival Analysis of Various Factors influencing Time to death

```{r}
# Extract df for sample survival analysis for variable HIV

HIV_survival_data <- rr_data_new %>%
  filter(`Date of Diagnosis` > ymd('2022-01-01') &
           `Date of Diagnosis` %--% `Date of Death or Transplant`/days() > 0 ) %>%
  filter(HIV %in% c('Pos', 'Neg')) %>% 
  mutate(`HIV status` = case_when(HIV == 'Neg'~ 0,
                                    HIV == 'Pos'~ 1),
           days = `Date of Diagnosis` %--% `Date of Death or Transplant`/days(),
           `death status` = case_when(Outcome == 'Dead' ~ 1,
                                      Outcome %in% c('Alive','Transplant') ~ 0)) %>% 
  select(days, `HIV status`, `death status`)
```

## Hazard Ratios

# Key Findings and Discussion

Kindly check the `.pptx` PowerPoint slides and report for details [here](https://github.com/nelsonudeme001/analysis_NGA_renal_registry/)
